Supporting Information: Thermal sensitivity across forest vertical profiles: patterns, mechanisms, and ecological implications

### 2022-08-03 journal submissions/2022_29_06/response-to-reviews.Rmd

---
title: ''
output: word_document
bibliography: Zotero_references.bib #This can be generated by exporting BibTex file from Zotero
csl: new-phytologist.csl

---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Dear Editor:

We are pleased to submit a revised version of our manuscript, "Thermal sensitivity across forest vertical profiles: patterns, mechanisms, and ecological implications" (NPH-TR-2021-38020), for consideration for publication in *New Phytologist*. We have addressed all points raised by the reviewers, as detailed in our responses to the reviewers. The most significant change was de-emphasizing the discussion of the potential for stronger negative effects of rising temperatures on understory trees, including removal of former Fig. 4b and modifying the text throughout to give a more consistent message (as requested by reviewer #2).

@@

Regarding production of the figures, below are some specific criteria:
<!--To ensure rapid handling of your manuscript we would be grateful if you could confirm, in the cover letter to the Editor, whether or not there are any important elements of the figures that need to be retained (eg. colours used, plant species drawn, layout and spacing, symbols, etc) and if you have specific instructions that you would like the illustrator to take on board.-->

- It will generally be fine to adjust colors, layout, and spacing on the figures.
- We would like to retain the background painting in Figure 1 (by the lead author)

@@

\newpage

# Response to Reviewer(s)' Comments to Author:

## Referee: 1

Comments to the Author

I am satisfied with the corrections and the responses to the comments. I like the new structure, the writting is very precise while covering broad areas and the paper very interesting.

@@

-Fig 3 caption: TLeaf and some other discreptances with rest of the paper abreviations or format (tair, gs)

**Fixed.**

-l304 Not sure about the structure of the sentence and causality. Gravity and hydraulic pathlength (due to height) could mitigate the effect of light alone ?

**The full sentence reads: "More recent work has extended the focus from light alone to *height* in the vertical profile, as *the latter* shapes differences in leaf structure and function that can partially mitigate the effects of gravity and hydraulic pathlength on leaves higher up in the canopy." Thus, we're saying that *height* shapes differences in leaf traits that could mitigate effects of gravity and hydraulic pathlength. We believe the sentence to be correct as written.**

@@

l.313 'information from adjacent mature leaves or apical meristems, including a “memory” of previous conditions in that location, ...'.  This was not in the previous version that I can see, and I don't see the basis for inclusion of this statement. As well, use of the term 'memory' seems awkward if not entirely wrong (even reading between quotes). Please revise.

**Revised as suggested.**

l.579 'implies that high 𝑇𝑎𝑖𝑟 should decrease 𝑔𝑠 of upper canopy leaves more than understory leaves'. The authors here are ignoring what they’ve said previously, that high Tair is associated with high VPD (l. 199-200), and that VPD is conclusively a much stronger driver of gs than Tleaf (Leuning 1995, Grossiord). I think the sentence should be corrected and include VPD.  I'm not yet aware of a good study that has decoupled VPD and Tleaf to prove that Tleaf intrinsically causes reduction in gs. So the authors cannot conclude this.

**We have removed the quoted sentence as it was redundant with that which had preceded it, but had confusingly couched the message of the preceding sentence as if it were a speculation rather than an empirical fact. The preceding sentence, which we have retained, read:**
*“…sun leaves in the upper canopy show a stronger decrease in* *g~s~* *in response to rising* *T~air~* *(driven by increased VPD) than lower-canopy shade leaves (Hernández et al., 2020; Carter et al., 2021)”.*
**That is an empirical result. The fact that the effect of *T~air~* on *g~s~* in that result is driven by VPD rather than somehow directly by *T~air~* itself is indicated by the fact that increasing temperature itself apparently *opens* stomata, rather than closing them. This is also germane to the reviewer’s final comment about decoupling VPD and *T~leaf~*. There are in fact six published studies that have decoupled VPD and *T~leaf~* [@schulzeRoleAirHumidity1974; @hallRegulationWaterLoss1975; @hallStomatalResponseEnvironment1975; @fredeenTemperatureHumidityEffects1999a; @mottStomatalResponsesHumidity2010; @urbanStomatalConductanceIncreases2017a], and they all suggest temperature per se causes an increase, not a reduction, in stomatal conductance. We have added the following sentence to the end of the paragraph to clarify this matter, citing three of the six studies due to limit on word count: "Available data also suggest that high Tair can directly drive stomatal opening, if vapor pressure is experimentally adjusted in tandem with Tair to keep VPD constant (Fredeen & Sage, 1999; Mott & Peak, 2010; Urban et al., 2017), though it appears that the negative effect of temperature on gs, mediated by VPD, generally overrides the positive effect of temperature per se in nature, causing gs to decline when the leaf warms."**

### 2022-08-04 vertical-thermal-review.Rmd
work on editorial requests
@NidhiVinod , FYI

**Authors:** 
Nidhi Vinod^1,2^,
Martijn Slot^3^,
Ian R. McGregor^4^,
Elsa M. Ordway^2,5^,
Marielle N. Smith^6^,
Tyeen C. Taylor^7^,
Lawren Sack^2^,
Thomas N. Buckley^8^,
Kristina J. Anderson-Teixeira^1,3^*

@@

3. Smithsonian Tropical Research Institute; Panama, Republic of Panama

@@

*T~leaf~* is strongly tied to, and usually within a few degrees of, *T~air~* [@drakeNoEvidenceHomeostatic2020;@rey-sanchez_spatial_2016], but can be substantially warmer or cooler under certain conditions [Note `r Notes_leafTbiophysics`,@doughtyAreTropicalForests2008;@rey-sanchez_spatial_2016;@vogelLeavesLowestHighest2009]. 
Deviation of *T~leaf~* from *T~air~* is influenced by other micrometeorological drivers and by leaf traits and stomatal conductance (*g~s~*; Fig. `r fig_leaf_T`; Note `r Notes_leafTbiophysics`), all of which vary across forest vertical gradients (Fig. `r fig_schematic`).
High radiation in the upper canopy implies that upper canopy leaves often have higher *T~leaf~* - *T~air~*, and thus are often warmer than understory leaves (Fig. `r fig_leaf_T`a).
However, higher wind speeds reduce *T~leaf~* - *T~air~* [Fig. `r fig_leaf_T`b, @niinemetsPhotosyntheticAcclimationSimultaneous2004; @bonan_ecological_2016]. 
In addition, adaptive leaf traits that increase *g~s~* and *g~b~*, such as small size and high stomatal density (Note `r Notes_leafTbiophysics`; sections 3, 4.1), mediate the direct effects of solar radiation on upper canopy leaves and result in higher rates of heat loss (Fig. `r fig_leaf_T`b-e). 
In contrast, in the lower canopy, greater RH (reduced VPD) and lower wind speeds (reduced *g~b~*) may limit leaf cooling via latent heat flux, increasing *T~leaf~* [@perezIncreasingHumidityThreatens2018; @tibbittsHumidityPlants1979].
Leaves in these normally-shaded locations can experience rapid light and temperature surges during sunflecks [>10°C, @waySunflecksTreesForests2012].

@@

Note `r Notes_leafTbiophysics`. Biophysical drivers of T~leaf~

>>>>

**Authors:**

Nidhi Vinod^1,2^

Martijn Slot^3^

Ian R. McGregor^4^ (Orcid ID: 0000-0002-5763-021X)

Elsa M. Ordway^2,5^

Marielle N. Smith^6^

Tyeen C. Taylor^7^

Lawren Sack^2^

Thomas N. Buckley^8^

Kristina J. Anderson-Teixeira^1,3^* ( Orcid ID : 0000-0001-8461-9713)

@@

3. Forest Global Earth Observatory, Smithsonian Tropical Research Institute, Apartado Postal 0843-03092, Panama City, Panama

@@

*T~leaf~* is strongly tied to, and usually within a few degrees of, *T~air~* [@drakeNoEvidenceHomeostatic2020;@rey-sanchez_spatial_2016], but can be substantially warmer or cooler under certain conditions [Notes `r Notes_leafTbiophysics`,@doughtyAreTropicalForests2008;@rey-sanchez_spatial_2016;@vogelLeavesLowestHighest2009]. 
Deviation of *T~leaf~* from *T~air~* is influenced by other micrometeorological drivers and by leaf traits and stomatal conductance (*g~s~*; Fig. `r fig_leaf_T`; Notes `r Notes_leafTbiophysics`), all of which vary across forest vertical gradients (Fig. `r fig_schematic`).
High radiation in the upper canopy implies that upper canopy leaves often have higher *T~leaf~* - *T~air~*, and thus are often warmer than understory leaves (Fig. `r fig_leaf_T`a).
However, higher wind speeds reduce *T~leaf~* - *T~air~* [Fig. `r fig_leaf_T`b, @niinemetsPhotosyntheticAcclimationSimultaneous2004; @bonan_ecological_2016]. 
In addition, adaptive leaf traits that increase *g~s~* and *g~b~*, such as small size and high stomatal density (Notes `r Notes_leafTbiophysics`; sections 3, 4.1), mediate the direct effects of solar radiation on upper canopy leaves and result in higher rates of heat loss (Fig. `r fig_leaf_T`b-e). 
In contrast, in the lower canopy, greater RH (reduced VPD) and lower wind speeds (reduced *g~b~*) may limit leaf cooling via latent heat flux, increasing *T~leaf~* [@perezIncreasingHumidityThreatens2018; @tibbittsHumidityPlants1979].
Leaves in these normally-shaded locations can experience rapid light and temperature surges during sunflecks [>10°C, @waySunflecksTreesForests2012].

@@

Notes `r Notes_leafTbiophysics`. Biophysical drivers of T~leaf~
